On the relation between metacognitive monitoring and domain knowledge

### 2020-08-10 ms.Rmd
Edits: Responses to reviewers

Metacognitive resolution is typically measured with metrics that are asserted to be independent of the participants' performance, skill, or knowledge in the primary task. For example, in a memory task, participants' memory abilities should not impact their metacognitive abilities, and vice versa. In an influential article @Nelsoncomparisoncurrentmeasures1984 recommended the Goodman-Kruskal gamma correlation [@GoodmanMeasuresAssociationCross1954] as a measure of resolution, partly because it is theoretically independent of task performance [see also @GonzalezMeasuringordinalassociation1996]. Other measures of resolution include Pearson's *r* [@NelsonNorms300generalinformation1980], the phi correlation [@KornellTransferMetacognitiveSkills2007], and signal detection theoretic measures [@benjamin2008measurement; @HighamNewimprovedgamma2018]. Although computationally different, these metrics measure the association between response accuracy and metacognitive rating in a way that is assumed to quantify metacognitive ability independent of task performance. However, it remains a possibility that metacognitive resolution is empirically related to task performance, with important implications to self-directed study.

@@

What could explain these mixed findings on the resolution-performance correlation? We suggest one answer: In some tasks, domain skill (task performance, or knowledge; we use these terms interchangeably) and common measures of resolution are inherently related. Although it has been repeatedly asserted that e.g. the gamma correlation does not depend on performance in the criterion test [@Griffineffectsdomainknowledge2009; @Nelsoncomparisoncurrentmeasures1984; @ThiedeElementaryschoolexperience2012], we suggest that the truth of this assertion depends on the method of testing [see also @GalvinTypetaskstheory2003; @Maniscalcosignaldetectiontheoretic2012; @schwartz1994methodological; @ThiedeDelayingStudentsMetacognitive1994].

@@

However, another similar study found that domain skill and resolution were positively related. In this study, participants studied five texts and rated how well they understood each text [@GriffinIndividualdifferencesrereading2008]. After all five texts, participants were tested on their comprehension of each text with 16-20 true/false questions. Resolution was measured with a Pearson's *r* correlation between comprehension ratings and proportion correct across tests for each participant (the authors reported that gamma correlations yielded similar results). In two experiments, resolution correlated positively with comprehension performance (when texts were read once, but not in a read-twice condition). These findings suggested the opposite from the previous study; that greater domain skill is associated with greater metacognitive resolution. Positive correlations in comprehension studies with multiple choice tests have also been found in other studies [@WileyImprovingmetacomprehensionaccuracy2016]. Further, manipulations that increase comprehension performance, such as re-reading texts, also increase resolution [the rereading effect; @RawsonrereadingeffectMetacomprehension2000].

These results suggest that the mixed findings regarding the resolution-performance relationship may at least partially be explained by features of the primary task: There was no relation between resolution and performance when performance was assessed with a free response method [@Griffineffectsdomainknowledge2009; but see @Thiedeimportancemonitoringselfregulation1999]. When a multiple choice method was used, a positive correlation emerged [@GriffinIndividualdifferencesrereading2008]. Similarly, in another multiple-choice experiment (@KellyMetacognitionemotionalface2011; Experiment 2), a positive resolution-performance correlation was observed. This observation led us to ask if there might be a common factor that leads to associative measures of metacognitive resolution to have some inherent relation to task performance under certain testing conditions, but not others.

@@

Our approach to answer this question was three-fold. First, we investigated a large previously unpublished dataset from six experiments that investigated middle school and university students' mathematics learning. These data allowed us to examine whether a robust performance-resolution correlation is found when multiple-choice testing is used. Second, we developed a hypothetical model of confidence ratings, and conducted simulations to investigate potential factors contributing to the relation between metacognitive resolution and domain skill. Third, we tested the predictions of this model in an experiment that directly contrasted free-response and multiple-choice testing.

@@

In what we call Experiment 1, we collated data from six experiments investigating mathematics learning. There were differences between experiments, such as testing medium (computer or paper) and age (middle school or undergraduate), but our focus here was not to investigate these factors' impact on mathematics performance or metacognition. Rather, we used this large dataset to examine the possible relationship between metacognitive resolution and (mathematics) performance, in a multiple-choice testing context. The common procedure was that participants completed mathematics tests consisting of 4-alternative multiple-choice problems, and rated their confidence in the accuracy of each of their answers. Because these experiments (1A-F) were so similar, we describe all six experiments' methods first, and then present the results together.

@@

```{r trivia-prep}
trivia <- read_rds("data/exp-trivia.rds") %>% 
  mutate(condition = factor(
    condition, 
    levels = c("free", "mafc"), 
    labels = c("Free", "2AFC")
  ))
tp <- read_rds("data/exp-trivia-participants.rds")
tsex <- sum(tp$Sex=="Female", na.rm = T)
tage <- range(tp$age, na.rm = T)
tfl <- sum(tp$`First Language`=="English", na.rm = T)
tmtt <- round(median(tp$time_taken/60), 0)
```

We recruited 100 participants through the Prolific (https://prolific.ac) online platform. `r tsex` participants were female, and age ranged from `r tage[1]` to `r tage[2]` years. We recruited participants currently residing in the US, and `r tfl` reported English as their first language. Participants were paid $3.64 for participating. All participants provided informed consent. The sample size was determined with a power simulation based on the results of the simulation study described above.

@@

We collected general knowledge questions from @TauberGeneralknowledgenorms2013 and from previous experiments conducted in our laboratory. Based on those previous studies, we selected 120 questions of varying degrees of difficulty, and found the most frequently endorsed commission errors to be used as the false response options in the 2AFC condition. We also screened the questions (and response alternatives) for potentially offensive or outdated material, and replaced unsuitable false response options with new ones. As an example, one question was "What was the name of the clown on the 'howdy doody' television show?". (Correct response: "Clarabell", incorrect alternative: "Bozo").

>>>>

Metacognitive resolution is typically measured with metrics that are asserted to be independent of the participants' performance, skill, or knowledge in the primary task. For example, in a memory task, participants' memory abilities should not impact their metacognitive abilities, and vice versa. That is, it is assumed that resolution is a distinct (meta-)cognitive ability to the abilities required in the primary task. In an influential article @Nelsoncomparisoncurrentmeasures1984 recommended the Goodman-Kruskal gamma correlation [@GoodmanMeasuresAssociationCross1954] as a measure of resolution, partly because it is theoretically independent of task performance [see also @GonzalezMeasuringordinalassociation1996]. Other measures of resolution include Pearson's *r* [@NelsonNorms300generalinformation1980], the phi correlation [@KornellTransferMetacognitiveSkills2007], and signal detection theoretic measures [@benjamin2008measurement; @HighamNewimprovedgamma2018]. Although computationally different, these metrics measure the association between response accuracy and metacognitive rating in a way that is assumed to quantify metacognitive ability independent of task performance. However, it remains a possibility that metacognitive resolution is empirically related to task performance, with important implications to self-directed study.

@@

What could explain these mixed findings on the resolution-performance correlation? We suggest one answer: In tasks using multiple-choice responding, domain skill (task performance, or knowledge; we use these terms interchangeably) and common measures of resolution are inherently related. Although it has been repeatedly asserted that e.g. the gamma correlation does not depend on performance in the criterion test [@Griffineffectsdomainknowledge2009; @Nelsoncomparisoncurrentmeasures1984; @ThiedeElementaryschoolexperience2012], we suggest that the truth of this assertion depends on the method of testing (multiple choice vs. free responses) [see also @GalvinTypetaskstheory2003; @Maniscalcosignaldetectiontheoretic2012; @schwartz1994methodological; @ThiedeDelayingStudentsMetacognitive1994].

@@

However, another similar study found that domain skill and resolution were positively related. In this study, participants studied five texts and rated how well they understood each text [@GriffinIndividualdifferencesrereading2008]. After all five texts, participants were tested on their comprehension of each text with 16-20 true/false questions. Resolution was measured with a Pearson's *r* correlation between comprehension ratings and proportion correct across tests for each participant (the authors reported that gamma correlations yielded similar results). In two experiments, resolution correlated positively with comprehension performance (when texts were read once, but not in a read-twice condition). These findings suggested the opposite from the previous study; that greater domain skill is associated with greater metacognitive resolution. Positive correlations in comprehension studies with multiple choice tests have also been found in other studies [@WileyImprovingmetacomprehensionaccuracy2016]. Further, manipulations that increase comprehension performance, such as re-reading texts, also increase resolution [the rereading effect; @RawsonrereadingeffectMetacomprehension2000], suggesting that performance and resolution may be related (either as separate abilities, or due to the measurement issues discussed herein).

These results suggest that the mixed findings regarding the resolution-performance relationship may at least partially be explained by features of the primary task: There was no relation between resolution and performance when performance was assessed with a free response method [@Griffineffectsdomainknowledge2009; but see @Thiedeimportancemonitoringselfregulation1999]. When a multiple choice method was used, a positive correlation emerged [@GriffinIndividualdifferencesrereading2008]. Similarly, in another multiple-choice experiment (@KellyMetacognitionemotionalface2011; Experiment 2), a positive resolution-performance correlation was observed. This observation led us to ask if there might be a common factor that leads to associative measures of metacognitive resolution to have some inherent relation to task performance under certain testing conditions, but not others. In the present work, we focus on confidence judgments after responding on the main task, but our main argument applies equally to e.g. judgments of learning before a response is provided.

@@

Our approach to answer this question was three-fold. First, we investigated a large previously unpublished dataset from six experiments that investigated middle school and university students' mathematics learning. These data allowed us to examine whether a robust performance-resolution correlation is found when multiple-choice testing is used. Based on prior literature, we hypothesized that a positive correlation would be found. Second, we developed a hypothetical model of confidence ratings, and conducted simulations to investigate potential factors contributing to the relation between metacognitive resolution and domain skill. Specifically, the simulation investigated the hypothesis that guessing may underlie the relation between domain skill and metacognitive resolution. Third, we tested the predictions of this model in an experiment that directly contrasted free-response and multiple-choice testing. In the second experiment, we hypothesized that measures of resolution in multiple-choice tasks would be i) smaller in magnitude than in free response tasks and ii) correlated to performance in the primary task (more strongly so than in free response tasks).

@@

In what we call Experiment 1, we collated data from six experiments investigating mathematics learning. There were differences between experiments, such as testing medium (computer or paper) and age (middle school or undergraduate), but our focus here was not to investigate these factors' impact on mathematics performance or metacognition. Rather, we used this large dataset to examine the possible relationship between metacognitive resolution and (mathematics) performance, in a multiple-choice testing context. The common procedure was that participants completed mathematics tests consisting of 4-alternative multiple-choice problems, and rated their confidence in the accuracy of each of their answers. If readers are interested in the potential differences between e.g. demographics, the data is openly available for reanalyses at https://osf.io/np3cu/. Because these experiments (1A-F) were so similar, we describe all six experiments' methods first, and then present the results together.

@@

```{r trivia-prep}
trivia <- read_rds("data/exp-trivia.rds") %>% 
  mutate(condition = factor(
    condition, 
    levels = c("free", "mafc"), 
    labels = c("Free", "2AFC")
  ))
tp <- read_rds("data/exp-trivia-participants.rds")
tsex <- sum(tp$Sex=="Female", na.rm = T)
tage <- range(tp$age, na.rm = T)
mage <- median(tp$age, na.rm = TRUE)
tfl <- sum(tp$`First Language`=="English", na.rm = T)
tmtt <- round(median(tp$time_taken/60), 0)
# quantile(tp$age, na.rm = TRUE) %>% round
```

We recruited 100 participants through the Prolific (https://prolific.ac) online platform. `r tsex` participants were female, and age ranged from `r tage[1]` to `r tage[2]` years (median = `r mage`). We recruited participants currently residing in the US, and `r tfl` reported English as their first language. Participants were paid $3.64 for participating, in line with the platform guidelines. All participants provided informed consent. The sample size was determined with a power simulation based on the results of the simulation study described above.

@@

We collected general knowledge questions from @TauberGeneralknowledgenorms2013 and from previous experiments conducted in our laboratory. Based on those previous studies, we selected 120 questions of varying degrees of difficulty, and found the most frequently endorsed commission errors to be used as the false response options in the 2AFC condition. We also screened the questions (and response alternatives) for potentially offensive or outdated material, and replaced unsuitable false response options with new ones. As an example, one question was "What was the name of the clown on the 'howdy doody' television show?". (Correct response: "Clarabell", incorrect alternative: "Bozo"). In this Experiment, we used general information questions rather than mathematics problems in order to generalize the findings from Experiment 1.
